---
title: "ExploratoryDataLecture"
output: html_document
---

These notes capture the discussion in Exploratory Data,
Week 4.

```{r load.training.data}
library(ggplot2)

# load the data
train.subject <- read.table('data/UCI HAR Dataset/train/subject_train.txt', header=F)
train.data <- read.table('data/UCI HAR Dataset/train/X_train.txt', header=F)
train.activity <- read.table('data/UCI HAR Dataset/train/y_train.txt', header=F)
```

```{r label.training.data}
# label the data.
labels = read.table('data/UCI HAR Dataset/features.txt', sep=' ')
labels <- make.names(labels[,2])
names(train.data) = make.names(labels)
names(train.subject)[1] = 'subject'
```

```{r convert.activity.to.factors}
# convert activity
labels = read.table('data/UCI HAR Dataset/activity_labels.txt')
train.activity[,2] = sapply(train.activity[,1], function(x) { labels[x, 2] })
names(train.activity) = c('activity', 'activity.level')
```

Now we want to look at the data for a single subject.

```{r select.one.subject}
# merge subject with data
train.sdata <- cbind(train.subject, train.activity, train.data)
# select one subject
subj1 <- subset(train.sdata, subject==1)
```

Plot acceleration with basic graphic system and activity-as-an-integer

```{r}
par(mfrow = c(3, 1), mar = c(5, 4, 1, 1))
#title('Mean Acceleration for Subject 1')
plot(subj1[,'tBodyAccMag.mean..'], x=subj1$activity, col=subj1$activity, ylab='Mean Acc')
plot(subj1[,'tBodyAccMag.max..'], x=subj1$activity, col=subj1$activity, ylab='Max Acc')
plot(subj1[,'tBodyAccJerkMag.max..'], x=subj1$activity, col=subj1$activity, ylab='Max Jerk')
```

Plot acceleration with basic graphic system and activity-as-an-factor

```{r}
par(mfrow = c(3, 1), mar = c(5, 4, 1, 1))
#title('Mean Acceleration for Subject 1')
plot(subj1[,'tBodyAccMag.mean..'], x=subj1$activity.level, col=subj1$activity, ylab='Mean Acc')
plot(subj1[,'tBodyAccMag.max..'], x=subj1$activity.level, col=subj1$activity, ylab='Max Acc')
plot(subj1[,'tBodyAccJerkMag.max..'], x=subj1$activity.level, col=subj1$activity, ylab='Max Jerk')
```

Perform Hierarchical Clustering on maximum acceleration.

```{r clustering}
hc <- hclust(dist(subj1[,c('activity','tBodyAccMag.max..')]))
plot(hc)
```

This shows a clear separation between idle activities on the left and active activities on the right.

Perform singular value decomposition.

```{r svd}
#names(subj1)
#names(subj1)[563]
#names(subj1)[564]
#svd1 = svd(scale(subj1[,-c(563,564)]))
#par(mfrow=c(1,2))
#plot(svd1$u[,1], col=subj1$activity, pch=19)
#plot(svd1$u[,2], col=subj1$activity, pch=19)
#head(svd1)
```



For comparison - plotting mean acceleration with ggplot system.

```{r mean.with.ggplot}
par(mfrow = c(2, 1), mar = c(5, 4, 1, 1))
ggplot(subj1, aes(x=activity.level, y= tBodyAccMag.mean..)) +
  geom_point()
ggplot(subj1, aes(x=activity.level, y= tBodyAccMag.max..)) +
  geom_point()
```

```{r}

# requires vector of size 206 MB
#distMatrix <- dist(s[,2:4])
#hc <- hclust(distMatrix)
#plot(hc)
#names(subj1)
```
