---
title: "Body Sensor Clustering Case Study"
output: html_document
---

This is notes based on coursera class https://class.coursera.org/exdata-008/lecture/71

Original data from: http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

## Data Processing

Load the data.

```{r}
#load("smartphones/data/UCI HAR Dataset")
# load the data
train.subject <- read.table('C:/users/bgiles/Documents/github/datasciencecoursera/smartphones/data/UCI HAR Dataset/train/subject_train.txt', header=F)
train.data <- read.table('C:/users/bgiles/Documents/github/datasciencecoursera/smartphones/data/UCI HAR Dataset/train/X_train.txt', header=F)
train.activity <- read.table('C:/users/bgiles/Documents/github/datasciencecoursera/smartphones/data/UCI HAR Dataset/train/y_train.txt', header=F)
```

Label the data.

```{r label.training.data}
# label the data.
labels = read.table('C:/users/bgiles/Documents/github/datasciencecoursera/smartphones/data/UCI HAR Dataset/features.txt', sep=' ')
labels <- make.names(labels[,2])
names(train.data) = make.names(labels)
names(train.subject)[1] = 'subject'

names(train.activity)[1] = 'activity'

# there's some way to read this from file...
train.activity[train.activity[,1] == 1 ,1] <- "WALKING"
train.activity[train.activity[,1] == 2, 1] <- "WALK UP"
train.activity[train.activity[,1] == 3, 1] <- "WALK DOWN"
train.activity[train.activity[,1] == 4, 1] <- "SITTING"
train.activity[train.activity[,1] == 5, 1] <- "STANDING"
train.activity[train.activity[,1] == 6, 1] <- "LAYING DOWN"

train.activity <- transform(train.activity, activity = factor(activity))

train <- cbind(train.subject, train.data, train.activity)
```

## plot average acceleration for first subject.

```{r}
par(mfrow = c(1, 2), mar = c(5, 4, 1, 1))
sub1 <- subset(train, train$subject == 1);

plot(sub1[, 2], col = sub1$activity, ylab = names(sub1)[2])
plot(sub1[, 3], col = sub1$activity, ylab = names(sub1)[3])
legend("bottomright", legend = unique(sub1$activity), col=unique(sub1$activity), pch = 1)
```

Perform clustering based on Euclidean distance

```
#source("myplclust.R")
#distanceMatrix <- dist(sub1[, 2:4])
#hclustering <- hclust(distanceMatrix)
#myplclust(hclustering, lab.col = unclass(sub1$activity))
```

(No clear clustering)

## plot max acceleration for first subject.

```{r}
par(mfrow = c(1, 2))
plot(sub1[,11], pch= 19, col = sub1$activity, ylab = names(sub1)[11])
plot(sub1[,12], pch= 19, col = sub1$activity, ylab = names(sub1)[12])
```

This shows that acceleration can probably be used to distinguish activities, at least active vs. stationary.

```{r}
#source("myplclust.R")
#distanceMatrix <- dist(sub1[, 11:13])
#hclustering <- hclust(distanceMatrix)
#myplclust(hclustering, lab.col = unclass(sub1$activity))
```

(shows clear clustering between active and stationary, no clear clustering within these two categories.)

## Singular Value Decomposition

Perform singular value decomposition to find most important factor.

```{r}
# drop first and last columns since they're subject and activity
svd1 = svd(scale(sub1[, -c(1, 563)]))
par(mfrow = c(1, 2))
plot(svd1$u[, 1], col = sub1$activity, pch = 19, ylab = names(sub1)[which.max(svd1$v[,2])])
plot(svd1$u[, 2], col = sub1$activity, pch = 19, ylab = names(sub1)[which.max(svd1$v[,4])])
```

First singular vector clearly separates motion vs. stationary. The second singluar vector is less clear - perhaps it separates out walking upstairs.

## Find maximum Contributor

```{r}
plot(svd1$v[, 2], phd = 19)

maxContrib <- which.max(svd1$v[,2])
names(sub1)[which.max(svd1$v[,2])]

# distanceMatrix <- dist(sub1[, c(11:13, maxContrib)])
# hclustering <- hclust(distanceMatrix)
# myplclust(hclustering, lab.col = unclass(sub1$activity))
```

(Plot shows clear separation between each motion (walk, walk up, walk down), but rest are still jumbled.)


```{r}
kClust <- kmeans(sub1[, -c(1,563)], centers = 6, nstart = 250)
table(kClust$cluster, sub1$activity)
```

Plot variable centers

```{r}
par(mfrow = c(2, 3))
plot(kClust$center[1, 1:10], pch = 19, ylab = "Center", xlab="1")
plot(kClust$center[2, 1:10], pch = 19, ylab = "Center", xlab="2")
plot(kClust$center[3, 1:10], pch = 19, ylab = "Center", xlab="3")
plot(kClust$center[4, 1:10], pch = 19, ylab = "Center", xlab="4")
plot(kClust$center[5, 1:10], pch = 19, ylab = "Center", xlab="5")
plot(kClust$center[6, 1:10], pch = 19, ylab = "Center", xlab="6")
